Estimation of Manning's Friction Coefficient using DART Buoys Data during \tohoku Tsunami

### 2013-12-23 paper/intro.tex
Revision of intro, redid wrapping

Tsunamis in the past decade have been responsible for some of the most 
deadly and costly natural disasters ever recorded. Coastal communities 
have faced this hazard by assessing the risk they pose, attempting 
to make informed decisions about the likelihood that such an event would 
occur with cost to life and property.  To do this, computational models of 
tsunami events are often employed to explore multiple different scenarios
to predict the water elevation caused by the tsunamis. The accurate prediction 
of water elevations requires accurate estimation of many model parameters that 
are either measured directly, defined empirically, or inferred from
sparsely collected data. Unfortunately, since tsunami events are relatively rare, 
there is an abundance of uncertainty in the input data for these computational 
models ranging from effects of the domain such as bathymetry and friction parameterizations, 
to the earthquake source.  In this article, we aim to estimate the Manning's n friction
coefficient that plays a key role in predicting the water elevation. 
Our approach is to quantify the uncertainty in the predicted water elevation 
and employ an inverse modeling approach to estimate the Manning's n coefficient 
using real observations of water elevation measured during the \tohoku tsunami as a case study.

Previous work looking into uncertainty in the input for tsunami simulations often has focused on the earthquake source but mainly as a way of constraining the various proposed earthquake scenarios only after the event (see for example \cite{MacInnes:2013cr}).  For instance, Sarri \emph{et al.} recently built a statistical surrogate of a landslide-generated tsunami computer model by using an emulator~\cite{Sarri2012}. The implemented emulator was the Gaussian process that requires using a combination of prior knowledge about the simulator and
appropriate choices of functions and parameters. Their work was presented as a proof-of-concept case study
where they performed basic statistical and sensitivity analysis. In this work we focus on the uncertainty in the friction parameterization used in tsunami simulations.  Changes to the friction parameterization has been shown in the past to lead to significant changes to inundation levels \cite{Myers:2001el,Jakeman:2010hk} but these previous studies were limited in the scope of the changes allowed.  In a recent work, Mayo et el. reformulated a statistical data assimilation method generally used in the estimation of model states to estimate Manning's n coefficient~\cite{Mayo2013}. They used a low-dimensional representations of Manning's n coefficients and recovered it by assimilating water elevation data.

In this work, we aim at making use of the recent developments in Uncertainty Quantification 
(UQ) methods that allows probing the sensitivity of realistic tsunami models 
to uncertain parameters, and inferring those parameters from specific observations. 
In particular, we implement an inverse modeling approach that relied on the Bayesian inference approach.  
We will also rely on Polynomial Chaos (PC) expansions to construct a 
faithful surrogate of the response of the large and complex geophysical 
model simulations.  The construction of the surrogate enables us to efficiently 
implement the Bayesian inference formalism as it dramatically reduces the cost of 
sampling the posterior distribution.

@@

In this work, we follow a similar procedure as in \cite{sraj:2013a,sraj:2013b}
to estimate Manning'n n coefficient of friction. 
%We use the widely used Manning's n law to provide a relationship between the friction coefficient and the roughness parameter $n$.  The parameter $n$ is also allowed to vary in space determined by the initial water depth to study effects of deep-water, near-shore, and on-shore sensitivities.  
To carry out the uncertainty quantification we have employed polynomial chaos expansions.  Here the quantity of interest is taken to be the surface elevation $\eta$ and will compared to available NOAA Deep-ocean Assessment and Reporting of Tsunamis (DART) buoys for the \tohoku tsunami.  The forward model runs are performed using \geoclaw following the setup and using one of the source models explored in \cite{MacInnes:2013cr}.

>>>>

Tsunamis in the past decade have been responsible for some of the most  deadly
and costly natural disasters ever recorded. Coastal communities  have faced this
hazard by assessing the risk they pose, attempting  to make informed decisions
about the likelihood that such an event would  occur with cost to life and
property.  To do this, computational models of  tsunami events are often
employed to explore multiple different scenarios to predict the water elevation
caused by the tsunamis. The accurate prediction  of water elevations requires
accurate estimation of many model parameters that  are either measured directly,
defined empirically, or inferred from sparsely collected data. Unfortunately,
since tsunami events are relatively rare,  there is an abundance of uncertainty
in the input data for these computational  models ranging from effects of the
domain such as bathymetry and friction parameterizations,  to the earthquake
source.  In this article, we aim to estimate the Manning's n friction
coefficient. Our approach quantifies the uncertainty in the predicted water
elevation  and employs an inverse modeling approach to estimate the Manning's
$n$ coefficient using observations of water elevation measured during an event.

Previous work looking into uncertainty in the input for tsunami simulations
often has focused on the earthquake source but primarily as a way of
constraining the various proposed earthquake scenarios only after the event (see
for example \cite{MacInnes:2013cr}).  A similar approach for landslide-generated
tsunami was presented in Sarri \emph{et al.}, building a statistical surrogate
using an emulator \cite{Sarri2012}.  The implemented emulator was the Gaussian
process that requires using a combination of prior knowledge about the simulator
and appropriate choices of functions and parameters. Their work was presented as
a proof-of-concept case study where they performed basic statistical and
sensitivity analysis. In this work we focus on the uncertainty in the friction
parameterization used in tsunami simulations.  Changes to the friction
parameterization has been shown in the past to lead to significant changes to
inundation levels \cite{Myers:2001el,Jakeman:2010hk} but these previous studies
were limited in the scope of the changes to the parameterization allowed.  In
recent work, Mayo et el. reformulated a statistical data assimilation method
generally used in the estimation of model states to estimate the Manning's n
coefficient~\cite{Mayo2013}. They used a low-dimensional representations of
Manning's n coefficients and recovered it by assimilating water elevation data.

\alert{Did the Talia et al. study use synthetic data?  Probably should mention
this if that was the case.}

The method proposed here utilizes recent developments in uncertainty
quantification (UQ) methods that allow non-intrusive probing of the sensitivity
of realistic tsunami  models to uncertain parameters, inferring those parameters
from specific  observations.  In particular, we implement an inverse modeling
approach that relies on the Bayesian inference approach.  We will also rely on
polynomial chaos (PC) expansions to construct a faithful surrogate of the
response of the large and complex geophysical model simulations.  THe
construction of the surrogate enables us to efficiently implement the Bayesian
inference formalism as it dramatically reduces the cost of sampling the
posterior distribution.

@@

In this work, we follow a similar procedure as in \cite{sraj:2013a,sraj:2013b}
to estimate Manning's $n$ coefficient of friction using the \tohoku tsunami
event as a case-study.  The quantity of interest is taken to be the surface
elevation $\eta$ and is compared to the available NOAA Deep-ocean Assessment and
Reporting Tsunamis (DART) buoys.  The forward model runs are performed using the
\geoclaw package following the setup in MacInnes \emph{et al.} including the
source models analyzed there \cite{MacInnes:2013cr}.
